---
title: '<span class="en">Day 3: Key Findings — Five Stories from Austrian Plant Diversity</span><span class="de">Tag 3: Schlüsselergebnisse — Fünf Geschichten der österreichischen Pflanzendiversität</span>'
subtitle: '<span class="en">A Discussion of the Most Compelling Results</span><span class="de">Eine Diskussion der überzeugendsten Ergebnisse</span>'
author: '<span class="en">Ecology Workshop</span><span class="de">Ökologie-Workshop</span>'
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6
)
library(ggplot2)
library(data.table)
library(spacc)
library(sf)
library(rnaturalearth)
library(hexify)

comp_dir  <- file.path(dirname(getwd()), "comp")
data_path <- file.path(dirname(getwd()), "data")
```

```{css, echo=FALSE}
/* ========== WORKSHOP THEME ========== */

body {
  font-family: 'Segoe UI', Tahoma, Geneva, Verdana, sans-serif;
  line-height: 1.7;
  background-color: #fafffe;
}

h1 { color: #2E7D32 !important; border-bottom: 3px solid #4CAF50; padding-bottom: 10px; }
h2 { color: #388E3C !important; margin-top: 2em; }
h3 { color: #43A047 !important; }

table { border-collapse: collapse; margin: 1em 0; width: 100%; border-radius: 8px; overflow: hidden; box-shadow: 0 2px 8px rgba(0,0,0,0.1); }
th { background-color: #4CAF50 !important; color: white !important; padding: 12px; text-align: left; }
td { padding: 10px 12px; border-bottom: 1px solid #ddd; }
tr:nth-child(even) { background-color: #f9f9f9; }
tr:hover { background-color: #E8F5E9; }

blockquote { border-left: 4px solid #FF9800; background-color: #FFF8E1; padding: 15px 20px; margin: 1em 0; border-radius: 0 8px 8px 0; }

hr { border: none; height: 3px; background: linear-gradient(to right, #4CAF50, #81C784, #C8E6C9, #81C784, #4CAF50); margin: 2em 0; border-radius: 2px; }

.tocify { border: none !important; border-radius: 12px !important; background-color: #f1f8e9 !important; }
.tocify-item.active { background-color: #4CAF50 !important; color: white !important; }

/* Hypothesis boxes */
.hypothesis-box {
  background: linear-gradient(135deg, #FFF3E0 0%, #FFE0B2 100%);
  border-left: 5px solid #FF9800;
  padding: 20px;
  margin: 1.5em 0;
  border-radius: 0 12px 12px 0;
}

/* Results boxes */
.results-box {
  background: linear-gradient(135deg, #E8F5E9 0%, #C8E6C9 100%);
  border-left: 5px solid #4CAF50;
  padding: 20px;
  margin: 1.5em 0;
  border-radius: 0 12px 12px 0;
}

/* Theory boxes */
.theory-box {
  background: linear-gradient(135deg, #E8F5E9 0%, #C8E6C9 100%);
  border-left: 5px solid #4CAF50;
  padding: 20px;
  margin: 1.5em 0;
  border-radius: 0 12px 12px 0;
}

/* Discussion boxes (new) */
.discussion-box {
  background: linear-gradient(135deg, #E3F2FD 0%, #BBDEFB 100%);
  border-left: 5px solid #1976D2;
  padding: 20px;
  margin: 1.5em 0;
  border-radius: 0 12px 12px 0;
}

/* Literature boxes */
.lit-box {
  background: linear-gradient(135deg, #F3E5F5 0%, #E1BEE7 100%);
  border-left: 5px solid #7B1FA2;
  padding: 20px;
  margin: 1.5em 0;
  border-radius: 0 12px 12px 0;
  font-size: 0.92em;
}

/* Math formula display fix */
.MathJax, .MathJax_Display, .mjx-chtml {
  overflow: visible !important;
  padding: 5px 0 !important;
}
p:has(.MathJax) {
  overflow: visible;
  line-height: 2.5;
}

/* ========== LANGUAGE TOGGLE ========== */
.lang-toggle {
  position: fixed; top: 10px; right: 20px; z-index: 9999;
  background: linear-gradient(135deg, #2E7D32, #4CAF50);
  padding: 8px 12px; border-radius: 25px;
  box-shadow: 0 4px 15px rgba(46, 125, 50, 0.3);
}
.lang-toggle button {
  background: transparent; border: none; color: #fff;
  padding: 6px 14px; cursor: pointer; font-weight: bold;
  border-radius: 20px; transition: all 0.3s; font-size: 14px;
}
.lang-toggle button.active { background: white; color: #2E7D32; }
.lang-toggle button:hover:not(.active) { background: rgba(255,255,255,0.2); }

.de { display: none; }
.en { display: inline; }
body.lang-de .de { display: inline; }
body.lang-de .en { display: none; }

div.de, p.de, li.de, h1.de, h2.de, h3.de, h4.de, section.de { display: none; }
div.en, p.en, li.en, h1.en, h2.en, h3.en, h4.en, section.en { display: block; }

body.lang-de div.de, body.lang-de p.de, body.lang-de li.de,
body.lang-de h1.de, body.lang-de h2.de, body.lang-de h3.de,
body.lang-de h4.de, body.lang-de section.de { display: block; }

body.lang-de div.en, body.lang-de p.en, body.lang-de li.en,
body.lang-de h1.en, body.lang-de h2.en, body.lang-de h3.en,
body.lang-de h4.en, body.lang-de section.en { display: none; }

body.lang-de #TOC .en { display: none; }
body.lang-de #TOC .de { display: inline; }
#TOC .de { display: none; }
#TOC .en { display: inline; }

@media print {
  .tocify, .lang-toggle { display: none; }
}

/* ========== MOBILE RESPONSIVE ========== */
@media (max-width: 768px) {
  body { font-size: 14px; line-height: 1.5; }
  h1 { font-size: 1.5rem; }
  h2 { font-size: 1.3rem; }
  h3 { font-size: 1.1rem; }
  .lang-toggle { top: 5px; right: 10px; padding: 5px 8px; }
  .lang-toggle button { padding: 4px 10px; font-size: 12px; }
  table { font-size: 12px; display: block; overflow-x: auto; white-space: nowrap; }
  th, td { padding: 6px 8px; }
  blockquote { padding: 10px 15px; margin: 0.5em 0; }
  pre { font-size: 11px; padding: 10px; }
  code { font-size: 11px; }
  img, .figure img { max-width: 100%; height: auto; }
}
```

```{js, echo=FALSE}
document.addEventListener('DOMContentLoaded', function() {
  var toggle = document.createElement('div');
  toggle.className = 'lang-toggle';
  toggle.innerHTML = '<button id="btn-en" class="active">EN</button><button id="btn-de">DE</button>';
  document.body.appendChild(toggle);

  var savedLang = localStorage.getItem('workshop-lang');
  var browserLang = navigator.language.substring(0, 2);
  var defaultLang = savedLang || (browserLang === 'de' ? 'de' : 'en');
  if (defaultLang === 'de') setLanguage('de');

  document.getElementById('btn-en').addEventListener('click', function() { setLanguage('en'); });
  document.getElementById('btn-de').addEventListener('click', function() { setLanguage('de'); });

  function setLanguage(lang) {
    document.body.classList.remove('lang-en', 'lang-de');
    document.body.classList.add('lang-' + lang);
    document.getElementById('btn-en').classList.toggle('active', lang === 'en');
    document.getElementById('btn-de').classList.toggle('active', lang === 'de');
    localStorage.setItem('workshop-lang', lang);
  }
});
```

---

```{r load-data}
header  <- fread(file.path(data_path, "austria_header.csv"))
species <- fread(file.path(data_path, "austria_species.csv"))

common_ids <- intersect(header$PlotObservationID, species$PlotObservationID)
header <- header[PlotObservationID %in% common_ids]
header[, decade := ifelse(Year < 1970, "pre-1970",
                          paste0(floor(Year / 10) * 10, "s"))]

n_plots   <- nrow(header)
n_records <- nrow(species)
n_species <- uniqueN(species$WFO_TAXON)
n_native  <- uniqueN(species[STATUS == "native"]$WFO_TAXON)
n_alien   <- uniqueN(species[STATUS == "neo"]$WFO_TAXON)
```

# <span class="en">Overview</span><span class="de">Überblick</span>

<div class="en">

This document distils the **five most compelling findings** from our comprehensive analysis of ~`r format(n_plots, big.mark = ",")` Austrian vegetation plots. Each finding is presented with its figure, followed by an extended discussion that connects the result to broader ecological theory, conservation practice, and open questions.

The five findings were selected because they each tell a distinct part of the Austrian biodiversity story:

| # | Finding | Why it matters |
|---|---------|----------------|
| 1 | **Where species live** — Hotspot maps | Spatial patterns reveal the geography of invasion |
| 2 | **How species accumulate** — Native vs. alien curves | The central hypothesis: different spatial strategies |
| 3 | **How much we're missing** — Invasion debt | The future of Austria's alien flora |
| 4 | **Why sites differ** — Turnover vs. nestedness | What drives Austria's high beta diversity |
| 5 | **Which habitats are vulnerable** — Habitat-level invasion | Where conservation efforts should focus |

</div>

<div class="de">

Dieses Dokument destilliert die **fünf überzeugendsten Ergebnisse** aus unserer umfassenden Analyse von ~`r format(n_plots, big.mark = ",")` österreichischen Vegetationsplots. Jedes Ergebnis wird mit seiner Abbildung präsentiert, gefolgt von einer erweiterten Diskussion, die das Ergebnis mit breiterer ökologischer Theorie, Naturschutzpraxis und offenen Fragen verbindet.

Die fünf Ergebnisse wurden ausgewählt, weil sie jeweils einen eigenständigen Teil der österreichischen Biodiversitätsgeschichte erzählen:

| # | Ergebnis | Warum es wichtig ist |
|---|----------|---------------------|
| 1 | **Wo Arten leben** — Hotspot-Karten | Räumliche Muster offenbaren die Geografie der Invasion |
| 2 | **Wie Arten akkumulieren** — Heimische vs. Alien-Kurven | Die zentrale Hypothese: verschiedene räumliche Strategien |
| 3 | **Was uns fehlt** — Invasionsschuld | Die Zukunft von Österreichs Alien-Flora |
| 4 | **Warum sich Standorte unterscheiden** — Turnover vs. Nestedness | Was Österreichs hohe Beta-Diversität antreibt |
| 5 | **Welche Habitate verwundbar sind** — Habitat-Invasionsmuster | Wo Naturschutzbemühungen fokussieren sollten |

</div>

---

# <span class="en">1. The Geography of Invasion: Hotspot Maps</span><span class="de">1. Die Geografie der Invasion: Hotspot-Karten</span>

## <span class="en">The Maps</span><span class="de">Die Karten</span>

```{r hotspot-data}
austria_sf <- ne_countries(scale = "medium", country = "Austria", returnclass = "sf")

plot_richness <- species[, .(
  n_native = uniqueN(WFO_TAXON[STATUS == "native"]),
  n_alien  = uniqueN(WFO_TAXON[STATUS == "neo"])
), by = PlotObservationID]

plot_richness[, total := n_native + n_alien]
plot_richness[, alien_pct := ifelse(total > 0, 100 * n_alien / total, 0)]

plot_rich_geo <- merge(plot_richness,
                       header[, .(PlotObservationID, Longitude, Latitude)],
                       by = "PlotObservationID")

grid <- hex_grid(area_km2 = 750)
hex_data <- hexify(plot_rich_geo, grid = grid, lon = "Longitude", lat = "Latitude")

hex_df <- as.data.frame(hex_data)
hex_summary <- as.data.table(hex_df)[, .(
  mean_native    = mean(n_native, na.rm = TRUE),
  mean_alien     = mean(n_alien, na.rm = TRUE),
  mean_alien_pct = mean(alien_pct, na.rm = TRUE),
  n_plots        = .N
), by = cell_id]

hex_summary <- hex_summary[n_plots >= 3]
hex_polys <- cell_to_sf(hex_summary$cell_id, grid = grid)
hex_polys <- merge(hex_polys, hex_summary, by = "cell_id")
```

```{r hotspot-native, fig.height=5.5}
ggplot() +
  geom_sf(data = austria_sf, fill = "grey95", color = "#333333", linewidth = 0.8) +
  geom_sf(data = hex_polys, aes(fill = mean_native), color = "white", linewidth = 0.3) +
  geom_sf(data = austria_sf, fill = NA, color = "#333333", linewidth = 1) +
  scale_fill_gradient(low = "white", high = "#2E7D32",
                      name = "Mean native\nrichness") +
  labs(title = "Native Species Richness Hotspots / Heimische Arten Hotspots",
       subtitle = "Mean species per plot, equal-area hexagonal grid (min. 3 plots)",
       x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 13) +
  coord_sf(xlim = c(9.3, 17.2), ylim = c(46.3, 49.1))
```

```{r hotspot-alien, fig.height=5.5}
ggplot() +
  geom_sf(data = austria_sf, fill = "grey95", color = "#333333", linewidth = 0.8) +
  geom_sf(data = hex_polys, aes(fill = mean_alien), color = "white", linewidth = 0.3) +
  geom_sf(data = austria_sf, fill = NA, color = "#333333", linewidth = 1) +
  scale_fill_gradient(low = "white", high = "#D32F2F",
                      name = "Mean alien\nrichness") +
  labs(title = "Alien Species Richness Hotspots / Alien-Arten Hotspots",
       subtitle = "Mean alien species per plot, equal-area hexagonal grid",
       x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 13) +
  coord_sf(xlim = c(9.3, 17.2), ylim = c(46.3, 49.1))
```

```{r hotspot-alien-pct, fig.height=5.5}
ggplot() +
  geom_sf(data = austria_sf, fill = "grey95", color = "#333333", linewidth = 0.8) +
  geom_sf(data = hex_polys, aes(fill = mean_alien_pct), color = "white", linewidth = 0.3) +
  geom_sf(data = austria_sf, fill = NA, color = "#333333", linewidth = 1) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       name = "Alien %") +
  labs(title = "Alien Species Proportion / Alien-Artenanteil",
       subtitle = "Mean % alien species per plot",
       x = "Longitude", y = "Latitude") +
  theme_minimal(base_size = 13) +
  coord_sf(xlim = c(9.3, 17.2), ylim = c(46.3, 49.1))
```

## <span class="en">Discussion</span><span class="de">Diskussion</span>

<div class="discussion-box">

<div class="en">

**What the maps show:**

These three hexagonal heatmaps reveal a striking spatial decoupling between native and alien plant richness across Austria. Native species richness peaks in the pre-Alpine foothills and montane zones, where overlapping vegetation belts create high habitat heterogeneity. Alien richness, by contrast, concentrates in the eastern lowlands, the Danube corridor, and around major urban centres.

**Why natives and aliens peak in different places:**

The native richness pattern reflects Austria's topographic complexity. Between 500 and 1500 m elevation, multiple vegetation zones converge: sub-montane deciduous forests, montane coniferous forests, and sub-alpine meadows. This elevational compression creates high local species richness through niche partitioning along temperature, moisture, and soil gradients (Rahbek 1995). The mid-elevation diversity peak is one of the most robust patterns in mountain ecology.

The alien pattern tells a different story. Alien species require **introduction pathways**: trade routes, transport infrastructure, gardens, agricultural imports. These pathways converge on cities and lowland corridors. The Danube valley serves as a continental-scale dispersal highway, connecting central Europe's botanical gardens, ports, and agricultural hinterlands. Vienna alone has been a gateway for ornamental and crop-associated aliens since the 18th century. The Pannonian Basin in eastern Austria adds warm, continental conditions that favour many thermophilic alien species from southern Europe and Asia.

**The proportion map is the most informative:**

While absolute richness maps can be confounded by sampling intensity (more plots in accessible areas), the alien *proportion* map controls for this by expressing aliens as a fraction of total diversity at each site. The clear east-west gradient confirms that the invasion pattern is not a sampling artefact. In the Pannonian lowlands, aliens can constitute 10--15% of local plot flora, while in the Alpine core they remain below 2%. This 5-fold difference in invasion intensity across a single country illustrates why national-scale analyses need spatial methods.

**Connections to invasion theory:**

The spatial concentration of aliens around human infrastructure supports the **propagule pressure** hypothesis (Lockwood et al. 2005): invasion success depends not just on species traits but on the rate at which individuals arrive. Urban areas, transport corridors, and agricultural land receive a continuous rain of alien propagules through trade goods, garden escapes, and contaminated soil. Alpine areas, though not intrinsically resistant, simply receive far fewer propagules.

This also connects to the **fluctuating resource** hypothesis (Davis et al. 2000): disturbed, nutrient-enriched lowland habitats experience regular resource pulses (tilling, fertilisation, construction) that create invasion windows. In contrast, the nutrient-poor, climatically harsh Alpine environment offers few such windows.

</div>

<div class="de">

**Was die Karten zeigen:**

Diese drei hexagonalen Heatmaps offenbaren eine auffällige räumliche Entkopplung zwischen heimischem und Alien-Pflanzenreichtum in Österreich. Der heimische Artenreichtum erreicht seinen Höhepunkt in den Voralpen und montanen Zonen, wo überlappende Vegetationsgürtel hohe Habitatheterogenität schaffen. Der Alien-Reichtum hingegen konzentriert sich in den östlichen Tieflagen, im Donaukorridor und um die großen Stadtzentren.

**Warum heimische und Alien-Arten an verschiedenen Orten ihre Höchstwerte erreichen:**

Das Muster des heimischen Reichtums spiegelt Österreichs topografische Komplexität wider. Zwischen 500 und 1500 m Höhe konvergieren mehrere Vegetationszonen: submontane Laubwälder, montane Nadelwälder und subalpine Wiesen. Diese Höhenkompression schafft hohen lokalen Artenreichtum durch Nischenaufteilung entlang von Temperatur-, Feuchtigkeits- und Bodengradienten (Rahbek 1995).

Das Alien-Muster erzählt eine andere Geschichte. Alien-Arten benötigen **Einführungswege**: Handelsrouten, Verkehrsinfrastruktur, Gärten, landwirtschaftliche Importe. Diese Wege konvergieren auf Städte und Tieflandkorridore. Das Donautal dient als kontinentale Ausbreitungsautobahn. Wien allein war seit dem 18. Jahrhundert ein Tor für ornamentale und kulturpflanzenassoziierte Aliens. Das Pannonische Becken in Ostösterreich bietet warme, kontinentale Bedingungen, die viele thermophile Alien-Arten aus Südeuropa und Asien begünstigen.

**Die Anteilskarte ist am informativsten:**

Während absolute Reichtumsarten durch Beprobungsintensität verfälscht werden können, kontrolliert die Alien-*Anteils*karte dies, indem sie Aliens als Bruchteil der Gesamtdiversität an jedem Standort ausdrückt. Der klare Ost-West-Gradient bestätigt, dass das Invasionsmuster kein Stichprobenartefakt ist. Im pannonischen Tiefland können Aliens 10--15% der lokalen Plot-Flora ausmachen, während sie im alpinen Kern unter 2% bleiben.

**Verbindungen zur Invasionstheorie:**

Die räumliche Konzentration von Aliens um menschliche Infrastruktur unterstützt die **Propagule-Pressure**-Hypothese (Lockwood et al. 2005): Invasionserfolg hängt nicht nur von Arteneigenschaften ab, sondern von der Rate, mit der Individuen ankommen. Die Verbindung zur **Hypothese der schwankenden Ressourcen** (Davis et al. 2000) liegt auf der Hand: gestörte, nährstoffangereicherte Tieflandhabitate erfahren regelmäßige Ressourcenpulse, die Invasionsfenster schaffen.

</div>

</div>

---

# <span class="en">2. The Central Finding: Aliens Accumulate Differently</span><span class="de">2. Das zentrale Ergebnis: Aliens akkumulieren anders</span>

## <span class="en">Native vs. Alien Accumulation Curves</span><span class="de">Heimische vs. Alien Akkumulationskurven</span>

```{r helper-functions}
# Normalized comparison plot for two spacc objects
plot_normalized <- function(..., title = "Species Accumulation Curves (normalized)",
                            ci = TRUE, ci_alpha = 0.3) {
  objs <- list(...)
  nms  <- names(objs)
  df_list <- lapply(seq_along(objs), function(i) {
    s <- summary(objs[[i]])
    total <- max(s$mean)
    data.frame(
      sites = s$sites, mean = s$mean / total,
      lower = s$lower / total, upper = s$upper / total, group = nms[i]
    )
  })
  df <- do.call(rbind, df_list)
  df$group <- factor(df$group, levels = nms)
  p <- ggplot(df, aes(x = sites, y = mean, color = group, fill = group))
  if (ci) p <- p + geom_ribbon(aes(ymin = lower, ymax = upper), alpha = ci_alpha, color = NA)
  p + geom_line(linewidth = 1) +
    scale_y_continuous(labels = scales::percent) +
    labs(x = "Sites sampled", y = "Proportion of total species",
         title = title, color = "Group", fill = "Group") +
    theme_minimal(base_size = 13)
}

# Normalized grouped plot
plot_grouped_normalized <- function(x, title = "Normalized Accumulation",
                                     ci = TRUE, ci_level = 0.95, ci_alpha = 0.3) {
  summaries <- summary(x, ci_level = ci_level)
  df_list <- lapply(names(summaries), function(g) {
    s <- summaries[[g]]
    if (length(s$sites) == 0) return(NULL)
    total <- max(s$mean)
    if (total == 0) return(NULL)
    data.frame(
      sites = s$sites, mean = s$mean / total,
      lower = s$lower / total, upper = s$upper / total, group = g
    )
  })
  df_list <- df_list[!vapply(df_list, is.null, logical(1))]
  df <- do.call(rbind, df_list)
  df$group <- factor(df$group)
  ggplot(df, aes(x = sites, y = mean, color = group, fill = group)) +
    { if (ci) geom_ribbon(aes(ymin = lower, ymax = upper), alpha = ci_alpha, color = NA) } +
    geom_line(linewidth = 1) +
    scale_y_continuous(labels = scales::percent) +
    scale_color_manual(values = c("native" = "#2E7D32", "neo" = "#D32F2F"),
                       labels = c("native" = "Native", "neo" = "Alien")) +
    scale_fill_manual(values = c("native" = "#2E7D32", "neo" = "#D32F2F"),
                      labels = c("native" = "Native", "neo" = "Alien")) +
    labs(x = "Sites sampled", y = "Proportion of total species",
         title = title, color = "Status", fill = "Status") +
    theme_minimal(base_size = 12)
}
```

```{r pooled-curves}
p <- readRDS(file.path(comp_dir, "pooled.rds"))
sac_native <- p$native; sac_alien <- p$alien; sac_all <- p$all
rm(p)

plot_normalized(Native = sac_native, Alien = sac_alien,
                title = "Pooled: Native vs Alien Accumulation (normalized)") +
  scale_color_manual(values = c("Native" = "#2E7D32", "Alien" = "#D32F2F")) +
  scale_fill_manual(values = c("Native" = "#2E7D32", "Alien" = "#D32F2F"))
```

```{r statistical-test}
compare <- readRDS(file.path(comp_dir, "compare.rds"))
print(compare)
plot(compare) +
  ggtitle("Permutation Test: Native vs Alien Curve Comparison") +
  theme_minimal(base_size = 13)
```

```{r grouped-status}
sac_grouped <- readRDS(file.path(comp_dir, "grouped_status.rds"))
plot_grouped_normalized(sac_grouped,
                        title = "Grouped by Status: Native vs Alien (1990s, normalized)") +
  scale_color_manual(values = c("native" = "#2E7D32", "neo" = "#D32F2F"),
                     labels = c("native" = "Native", "neo" = "Alien")) +
  scale_fill_manual(values = c("native" = "#2E7D32", "neo" = "#D32F2F"),
                    labels = c("native" = "Native", "neo" = "Alien"))
```

## <span class="en">Discussion</span><span class="de">Diskussion</span>

<div class="discussion-box">

<div class="en">

**The core result:**

When we normalize both curves to their respective totals, the alien curve lies consistently below the native curve. This means that as we walk across the Austrian landscape sampling plots in spatial order, we discover a *smaller fraction* of the alien flora at each step compared to the native flora. The permutation test confirms this difference is statistically significant (p < 0.05), and the grouped analysis (where both groups are accumulated along the exact same spatial paths) eliminates any confounding from different walk trajectories.

**What does a lower normalized curve mean biologically?**

A lower normalized curve indicates higher **spatial beta diversity**: the species composition changes more between distant sites. For natives, many widespread generalist species (common grasses, widespread forest herbs) appear early in the spatial walk and are shared across much of the landscape. Once you have sampled a few hundred plots from different regions, you have already encountered 50-60% of native species. For aliens, the story differs: each region harbours a partly different set of aliens, reflecting their distinct introduction histories and habitat preferences. A spatial walk through the Vienna Basin encounters different aliens than one through the Inn Valley or the Klagenfurt Basin.

**The multi-focus introduction model:**

This pattern is best explained by what we call the **multi-focus introduction model**. Native species have had millennia to reach equilibrium distributions governed by climate and soil. Alien species, by contrast, arrived at specific introduction points (ports, cities, botanical gardens, agricultural areas) at different times and from different source regions. Each introduction point acts as a separate focus of establishment and local spread:

- **Vienna and eastern Austria** received Pannonian and Eurasian steppe aliens through trade along the Danube
- **Innsbruck and western Austria** received Mediterranean and Alpine-pass aliens through the Brenner corridor
- **Graz and southern Austria** received Balkan and sub-Mediterranean aliens through southern trade routes

Because each region received a partly different alien species pool, more of the landscape must be traversed to encounter the full complement. This spatial heterogeneity in alien composition is the signature of a flora still in the process of range expansion.

**Temporal robustness:**

The decade-by-decade analysis (shown in the full results) confirms this pattern is stable across all sampling periods from the 1970s to the 2010s. The alien curve shape does not converge toward the native shape over time, suggesting that alien species are not homogenizing across the landscape, at least not yet. This challenges the expectation that with time, aliens would spread everywhere and erase spatial differences.

**Implications for monitoring:**

This finding has practical consequences for biodiversity monitoring. If a monitoring programme samples only a single region, it will systematically underestimate alien diversity more than native diversity. To achieve the same proportional coverage of alien species as natives, a monitoring network must span more of Austria's geographic and environmental extent. This argues for spatially distributed monitoring designs over intensive local surveys.

</div>

<div class="de">

**Das Kernergebnis:**

Wenn wir beide Kurven auf ihre jeweiligen Gesamtwerte normalisieren, liegt die Alien-Kurve konsistent unter der heimischen Kurve. Das bedeutet, dass wir beim Durchqueren der österreichischen Landschaft in räumlicher Reihenfolge bei jedem Schritt einen *kleineren Anteil* der Alien-Flora entdecken als der heimischen Flora. Der Permutationstest bestätigt, dass dieser Unterschied statistisch signifikant ist (p < 0,05), und die gruppierte Analyse (bei der beide Gruppen entlang exakt derselben räumlichen Pfade akkumuliert werden) eliminiert jeden Störeffekt verschiedener Walk-Trajektorien.

**Was bedeutet eine niedrigere normalisierte Kurve biologisch?**

Eine niedrigere normalisierte Kurve zeigt höhere **räumliche Beta-Diversität** an: Die Artenzusammensetzung ändert sich stärker zwischen entfernten Standorten. Bei Einheimischen erscheinen viele weit verbreitete Generalistenarten (häufige Gräser, weit verbreitete Waldkräuter) früh im räumlichen Walk und werden über weite Teile der Landschaft geteilt. Bei Aliens unterscheidet sich die Geschichte: Jede Region beherbergt eine teils unterschiedliche Menge an Aliens, was ihre verschiedenen Einführungsgeschichten und Habitatpräferenzen widerspiegelt.

**Das Multi-Fokus-Einführungsmodell:**

Dieses Muster wird am besten durch das **Multi-Fokus-Einführungsmodell** erklärt. Heimische Arten hatten Jahrtausende, um Gleichgewichtsverteilungen zu erreichen. Alien-Arten hingegen kamen an spezifischen Einführungspunkten (Häfen, Städte, Botanische Gärten, Agrargebiete) zu verschiedenen Zeiten und aus verschiedenen Quellregionen an:

- **Wien und Ostösterreich** erhielten pannonische und eurasische Steppenaliens durch den Donauhandel
- **Innsbruck und Westösterreich** erhielten mediterrane und Alpenpass-Aliens durch den Brennerkorridor
- **Graz und Südösterreich** erhielten balkanische und submediterrane Aliens durch südliche Handelsrouten

Da jede Region einen teils unterschiedlichen Alien-Artenpool erhielt, muss mehr der Landschaft durchquert werden, um das vollständige Komplement zu finden.

**Zeitliche Robustheit:**

Die jahrzehntweise Analyse bestätigt, dass dieses Muster über alle Beprobungszeiträume von den 1970ern bis zu den 2010ern stabil ist. Die Alien-Kurvenform konvergiert nicht zur heimischen Form, was darauf hindeutet, dass Alien-Arten die Landschaft (noch) nicht homogenisieren.

**Implikationen für das Monitoring:**

Wenn ein Monitoringprogramm nur eine einzelne Region beprobt, wird es die Alien-Diversität systematisch stärker unterschätzen als die heimische. Um denselben proportionalen Erfassungsgrad für Alien-Arten zu erreichen, muss ein Monitoringnetzwerk mehr von Österreichs geografischer Ausdehnung umfassen.

</div>

</div>

---

# <span class="en">3. The Invasion Debt: How Many Aliens Are We Missing?</span><span class="de">3. Die Invasionsschuld: Wie viele Aliens fehlen uns?</span>

## <span class="en">Michaelis-Menten Extrapolation</span><span class="de">Michaelis-Menten-Extrapolation</span>

<div class="theory-box">

<div class="en">

The **Michaelis-Menten model** fits a saturating curve to the species accumulation data:

$$S(n) = \frac{S_{max} \cdot n}{K_m + n}$$

where $S_{max}$ is the estimated total species richness (the asymptote) and $K_m$ is the half-saturation constant. The gap between the observed species count and the estimated $S_{max}$ represents **dark diversity** — species present in the region but not yet detected (Partel et al. 2011).

</div>

<div class="de">

Das **Michaelis-Menten-Modell** passt eine sättigende Kurve an die Artenakkumulationsdaten an:

$$S(n) = \frac{S_{max} \cdot n}{K_m + n}$$

wobei $S_{max}$ der geschätzte Gesamtartenreichtum (die Asymptote) und $K_m$ die Halbsättigungskonstante ist. Die Lücke zwischen der beobachteten Artenzahl und dem geschätzten $S_{max}$ repräsentiert **dunkle Diversität** — Arten, die in der Region vorhanden, aber noch nicht nachgewiesen sind (Partel et al. 2011).

</div>

</div>

```{r extrapolation}
e <- readRDS(file.path(comp_dir, "extrapolate.rds"))
fit_native <- e$native; fit_alien <- e$alien
rm(e)
```

```{r extrap-native, fig.height=5}
cat("Native species extrapolation:\n")
print(fit_native)
plot(fit_native) + ggtitle("Native Species: Observed vs Extrapolated")
```

```{r extrap-alien, fig.height=5}
cat("\nAlien species extrapolation:\n")
print(fit_alien)
plot(fit_alien) + ggtitle("Alien Species: Observed vs Extrapolated")
```

## <span class="en">Discussion</span><span class="de">Diskussion</span>

<div class="discussion-box">

<div class="en">

**The numbers:**

- **Native species:** ~2,378 observed out of ~2,525 estimated = **~94% completeness**
- **Alien species:** ~323 observed out of ~411 estimated = **~79% completeness**

The 15-percentage-point gap in sampling completeness between natives and aliens is one of the most striking results from the entire analysis. It is not simply that we have fewer alien species; we are also proportionally *further from knowing them all*.

**Why is the alien flora less complete?**

Three complementary mechanisms explain this gap:

1. **Ongoing introductions.** New alien species continue arriving in Austria through international trade, horticulture, and climate-driven range shifts. The global accumulation of alien plant species shows no sign of saturation (Seebens et al. 2017). Each year brings new garden escapes, contaminants in seed lots, and hitchhikers on imported goods. The "missing" ~88 alien species are not necessarily undiscovered in nature; some may not have arrived yet but will under current trade regimes.

2. **Dispersal lag.** Many alien species already present in Austria have not yet spread to all suitable habitats. A species introduced to a Viennese botanical garden in the 1990s may take decades to colonize the Danube floodplains, the railway embankments of Upper Austria, or the ruderal sites of Carinthia. During this spreading phase, the species exists in Austria but is detectable only at a few sites. With 52,000 plots spread unevenly across the country, many such early-stage colonizers are missed.

3. **Detection difficulty.** Some alien species are taxonomically challenging (critical groups where identification requires specialist knowledge) or occupy habitats that are under-sampled (urban wastelands, industrial sites, private gardens). These species are present but effectively invisible to standard vegetation surveys.

**The concept of invasion debt:**

Essl et al. (2011) introduced the concept of **invasion debt** by analogy with extinction debt: just as habitat loss creates a time-delayed wave of extinctions, past introductions create a time-delayed wave of invasions. Species that have been introduced but not yet spread represent a "debt" that will be "paid" through future range expansion, even if no new species are introduced.

For Austria, our estimate of ~88 "missing" alien species represents the upper bound of this debt. Some of these species may already be present in under-sampled areas (they are dark diversity), while others may arrive through future introductions. Either way, Austria's alien flora is likely to grow by approximately 25% from its current observed level, even under a hypothetical scenario of zero new introductions.

**What this means for conservation planning:**

Conservation strategies that use current alien species lists as static baselines will underestimate future management needs. Monitoring programmes should plan for a growing alien flora and allocate resources for early detection surveys in high-risk habitats (urban fringes, transport corridors, river valleys). The invasion debt concept also means that the success of biosecurity measures cannot be judged by current invasion levels alone; the effects of today's introductions may not manifest for decades.

**A cautionary note on the model:**

The Michaelis-Menten estimator assumes that the accumulation curve follows a specific functional form. If the true relationship between sampling effort and species discovery is more complex (e.g., if there is a long tail of very rare species), the asymptote may be underestimated. Chao-type estimators and Good-Turing corrections could provide alternative estimates. The 94% and 79% figures should be treated as reasonable approximations rather than precise values.

</div>

<div class="de">

**Die Zahlen:**

- **Heimische Arten:** ~2.378 beobachtet von ~2.525 geschätzt = **~94% Vollständigkeit**
- **Alien-Arten:** ~323 beobachtet von ~411 geschätzt = **~79% Vollständigkeit**

Die 15-Prozentpunkt-Lücke in der Beprobungsvollständigkeit zwischen Einheimischen und Aliens ist eines der auffälligsten Ergebnisse der gesamten Analyse.

**Warum ist die Alien-Flora weniger vollständig?**

Drei komplementäre Mechanismen erklären diese Lücke:

1. **Fortlaufende Einführungen.** Neue Alien-Arten kommen weiterhin nach Österreich durch internationalen Handel, Gartenbau und klimabedingte Arealverschiebungen. Die globale Akkumulation von Alien-Pflanzenarten zeigt kein Zeichen der Sättigung (Seebens et al. 2017).

2. **Ausbreitungsverzögerung.** Viele bereits in Österreich vorhandene Alien-Arten haben sich noch nicht auf alle geeigneten Habitate ausgebreitet. Eine in den 1990ern in einem Wiener Botanischen Garten eingeführte Art kann Jahrzehnte brauchen, um die Donauauen oder die Ruderalstandorte Kärntens zu besiedeln.

3. **Nachweisschwierigkeit.** Einige Alien-Arten sind taxonomisch herausfordernd oder besetzen unterbeprobte Habitate (urbane Brachen, Industriegelände, Privatgärten).

**Das Konzept der Invasionsschuld:**

Essl et al. (2011) führten das Konzept der **Invasionsschuld** ein: So wie Habitatverlust eine zeitverzögerte Aussterbewelle erzeugt, erzeugen vergangene Einführungen eine zeitverzögerte Invasionswelle. Für Österreich repräsentieren unsere geschätzten ~88 "fehlenden" Alien-Arten die Obergrenze dieser Schuld. Österreichs Alien-Flora wird voraussichtlich um etwa 25% von ihrem aktuellen Niveau wachsen, selbst ohne neue Einführungen.

**Was das für die Naturschutzplanung bedeutet:**

Naturschutzstrategien, die aktuelle Alien-Artenlisten als statische Baselines verwenden, werden den zukünftigen Managementbedarf unterschätzen. Monitoringprogramme sollten eine wachsende Alien-Flora einplanen und Ressourcen für Früherkennungserhebungen in Hochrisikohabitaten bereitstellen.

**Modellvorbehalt:**

Der Michaelis-Menten-Schätzer nimmt an, dass die Akkumulationskurve einer bestimmten Funktionsform folgt. Die Werte von 94% und 79% sollten als vernünftige Näherungen betrachtet werden, nicht als exakte Werte.

</div>

</div>

---

# <span class="en">4. Turnover, Not Nestedness: What Drives Austrian Beta Diversity</span><span class="de">4. Turnover, nicht Nestedness: Was Österreichs Beta-Diversität antreibt</span>

## <span class="en">Decomposing Beta Diversity</span><span class="de">Zerlegung der Beta-Diversität</span>

<div class="theory-box">

<div class="en">

Baselga (2010) showed that total beta diversity can be decomposed into two components:

$$\beta_{Sorensen} = \beta_{Turnover} + \beta_{Nestedness}$$

- **Turnover ($\beta_{sim}$):** Species are replaced by different species between sites
- **Nestedness ($\beta_{nes}$):** Species-poor sites are subsets of species-rich sites

The balance between these components reveals the *process* generating diversity differences: environmental filtering (turnover) versus differential species loss (nestedness).

</div>

<div class="de">

Baselga (2010) zeigte, dass die Gesamt-Beta-Diversität in zwei Komponenten zerlegt werden kann:

$$\beta_{Sorensen} = \beta_{Turnover} + \beta_{Nestedness}$$

- **Turnover ($\beta_{sim}$):** Arten werden zwischen Standorten durch andere Arten ersetzt
- **Nestedness ($\beta_{nes}$):** Artenarme Standorte sind Teilmengen artenreicher Standorte

Das Gleichgewicht zwischen diesen Komponenten verrät den *Prozess*, der Diversitätsunterschiede erzeugt: Umweltfilterung (Turnover) versus differenzieller Artenverlust (Nestedness).

</div>

</div>

```{r beta-diversity}
beta <- readRDS(file.path(comp_dir, "beta.rds"))
print(beta)
plot(beta) +
  ggtitle("Beta Diversity Accumulation: Turnover vs Nestedness (1990s)") +
  theme_minimal(base_size = 13)
```

## <span class="en">Discussion</span><span class="de">Diskussion</span>

<div class="discussion-box">

<div class="en">

**Turnover dominates overwhelmingly.**

As the spatial sampling window expands across Austria, the beta diversity decomposition reveals that **species turnover** accounts for the vast majority of compositional differences between sites. Nestedness contributes comparatively little.

**What this tells us about Austria's ecology:**

Austria spans an extraordinary elevation gradient from 115 m (Neusiedler See, Burgenland) to 3,798 m (Grossglockner, Carinthia) within a country only ~575 km wide. This 3,683 m elevation range creates a near-continuous series of vegetation belts:

1. **Pannonian steppe grasslands** (115--300 m): thermophilic, drought-tolerant species like *Festuca rupicola*, *Stipa* spp.
2. **Colline deciduous forests** (300--700 m): *Quercus*, *Carpinus*, *Tilia* with rich herb layers
3. **Montane conifer-broadleaf transition** (700--1200 m): *Fagus sylvatica* gives way to *Picea abies*
4. **Subalpine coniferous forests** (1200--1800 m): *Picea*, *Larix decidua*, *Pinus cembra*
5. **Alpine meadows and dwarf shrubs** (1800--2500 m): *Carex curvula*, *Loiseleuria procumbens*, *Salix* spp.
6. **Subnival and nival zone** (>2500 m): *Ranunculus glacialis*, *Saxifraga* spp., cushion plants

These zones host genuinely **different species**, not just nested subsets of the same flora. A lowland *Quercus*-*Carpinus* forest and a high-alpine *Carex curvula* grassland share virtually no species. This species replacement along elevation gradients is the classic signature of environmental filtering through temperature, growing season length, snow cover duration, and soil development.

**Why nestedness is low:**

Low nestedness means that species-poor sites (e.g., high alpine areas with 15--30 species per plot) are not simply impoverished versions of species-rich lowland sites (40--80 species per plot). Instead, they host entirely different species. An alpine meadow lacks lowland species not because it has "lost" them but because the environmental conditions exclude them entirely and support a distinct, cold-adapted community.

This contrasts with archipelago systems where nestedness dominates: small islands contain subsets of the species found on large islands, reflecting area-dependent extinction. Austria's flora is shaped by environmental gradients, not by area-dependent processes.

**Conservation implications:**

The dominance of turnover has a direct and important consequence for conservation planning. In a turnover-dominated system, **no single protected area can capture the full flora**. A large national park in the Alps protects alpine specialists but misses Pannonian steppe species entirely. Conversely, a lowland nature reserve near the Neusiedler See captures thermophilic specialists absent from any mountain reserve.

This supports the "**Several Small**" position in the SLOSS debate (Tjorve 2010) for Austria: a distributed network of smaller reserves spanning different elevations and biogeographic regions will capture more total diversity than a single large reserve of equivalent total area. Austria's existing protected area network partially reflects this, with reserves spanning from the Neusiedler See-Seewinkel National Park (lowland Pannonian) through the Kalkalpen National Park (montane) to the Hohe Tauern National Park (alpine-nival). Our analysis provides quantitative support for maintaining and expanding this distributed approach.

**Connection to the invasion story:**

The turnover result also contextualizes the alien species pattern. Because turnover dominates, the alien flora at each location reflects the specific combination of local environmental conditions and introduction history. Different lowland regions receive different aliens (Pannonian vs. continental vs. sub-Mediterranean), and these aliens barely penetrate the alpine zone. The invasion is not a uniform wave spreading outward; it is a mosaic of local invasions, each shaped by the local environment and human land use.

</div>

<div class="de">

**Turnover dominiert überwältigend.**

Wenn sich das räumliche Beprobungsfenster über Österreich erweitert, zeigt die Beta-Diversitäts-Zerlegung, dass **Artenersatz** den Großteil der kompositorischen Unterschiede zwischen Standorten ausmacht. Nestedness trägt vergleichsweise wenig bei.

**Was das über Österreichs Ökologie aussagt:**

Österreich umfasst einen außergewöhnlichen Höhengradienten von 115 m (Neusiedler See) bis 3.798 m (Großglockner) innerhalb eines nur ~575 km breiten Landes. Dieser 3.683-m-Höhenbereich schafft eine nahezu kontinuierliche Reihe von Vegetationsgürteln:

1. **Pannonische Steppen** (115--300 m): thermophile, trockenheitstolerante Arten
2. **Kolline Laubwälder** (300--700 m): *Quercus*, *Carpinus*, *Tilia*
3. **Montaner Übergang** (700--1200 m): *Fagus sylvatica* weicht *Picea abies*
4. **Subalpine Nadelwälder** (1200--1800 m): *Picea*, *Larix decidua*, *Pinus cembra*
5. **Alpine Wiesen** (1800--2500 m): *Carex curvula*, *Loiseleuria procumbens*
6. **Subnival-nival** (>2500 m): *Ranunculus glacialis*, *Saxifraga* spp.

Diese Zonen beherbergen **tatsächlich verschiedene Arten**, nicht einfach verschachtelte Teilmengen derselben Flora. Ein Tiefland-*Quercus*-*Carpinus*-Wald und ein hochalpines *Carex curvula*-Grasland teilen praktisch keine Arten.

**Warum Nestedness niedrig ist:**

Geringe Nestedness bedeutet, dass artenarme Standorte (z.B. hochalpine Gebiete mit 15--30 Arten pro Plot) nicht einfach verarmte Versionen artenreicher Tieflandstandorte (40--80 Arten) sind. Stattdessen beherbergen sie völlig andere Arten. Eine alpine Wiese fehlt Tieflandarten nicht, weil sie sie "verloren" hat, sondern weil die Umweltbedingungen sie vollständig ausschließen.

**Naturschutzimplikationen:**

Die Dominanz des Turnover hat eine direkte Konsequenz für die Naturschutzplanung. In einem turnover-dominierten System **kann kein einzelnes Schutzgebiet die gesamte Flora erfassen**. Dies unterstützt die "**Several Small**"-Position in der SLOSS-Debatte (Tjorve 2010) für Österreich: Ein verteiltes Netzwerk kleinerer Reservate, das verschiedene Höhenlagen und biogeografische Regionen umfasst, wird mehr Gesamtdiversität erfassen als ein einzelnes großes Reservat äquivalenter Gesamtfläche.

**Verbindung zur Invasionsgeschichte:**

Das Turnover-Ergebnis kontextualisiert auch das Alien-Artenmuster. Weil Turnover dominiert, spiegelt die Alien-Flora an jedem Standort die spezifische Kombination aus lokalen Umweltbedingungen und Einführungsgeschichte wider. Die Invasion ist keine gleichmäßige Welle; sie ist ein Mosaik lokaler Invasionen.

</div>

</div>

---

# <span class="en">5. Habitat Vulnerability: Where Invasions Concentrate</span><span class="de">5. Habitat-Verwundbarkeit: Wo sich Invasionen konzentrieren</span>

## <span class="en">Alien Species Share by Habitat</span><span class="de">Alien-Artenanteil nach Habitat</span>

```{r habitat-alien-share}
hab_alien <- header[, .(PlotObservationID, Eunis_lvl1)]
hab_alien <- merge(hab_alien,
                   species[, .(n_native = sum(STATUS == "native"),
                               n_alien  = sum(STATUS == "neo")),
                           by = PlotObservationID],
                   by = "PlotObservationID")
keep_habs <- names(which(table(header$Eunis_lvl1) >= 100))
hab_summary <- hab_alien[Eunis_lvl1 %in% keep_habs,
                         .(mean_native = round(mean(n_native), 1),
                           mean_alien  = round(mean(n_alien), 1),
                           alien_pct   = round(mean(n_alien / (n_native + n_alien)) * 100, 1)),
                         by = Eunis_lvl1]
setorder(hab_summary, -alien_pct)
knitr::kable(hab_summary,
             col.names = c("EUNIS Habitat", "Mean Native / Mittl. Heimisch",
                           "Mean Alien / Mittl. Alien", "Alien % / Alien %"),
             align = "lrrr")
```

## <span class="en">Accumulation Curves by Habitat</span><span class="de">Akkumulationskurven nach Habitat</span>

```{r habitat-curves}
hab_results <- readRDS(file.path(comp_dir, "habitat_curves.rds"))
hab_combined <- do.call(c, hab_results)
plot(hab_combined) +
  ggtitle("Species Accumulation by EUNIS Habitat (level 1)") +
  theme_minimal(base_size = 13)
```

```{r habitat-native-alien, fig.height=12}
hab_na_results <- readRDS(file.path(comp_dir, "habitat_native_alien.rds"))

plots_hab <- lapply(keep_habs, function(h) {
  if (is.null(hab_na_results[[h]])) return(NULL)
  plot_normalized(Native = hab_na_results[[h]]$native,
                  Alien  = hab_na_results[[h]]$alien,
                  title = sprintf("Habitat %s: Native vs Alien (normalized)", h)) +
    scale_color_manual(values = c("Native" = "#2E7D32", "Alien" = "#D32F2F")) +
    scale_fill_manual(values = c("Native" = "#2E7D32", "Alien" = "#D32F2F"))
})
plots_hab <- plots_hab[!vapply(plots_hab, is.null, logical(1))]
gridExtra::grid.arrange(grobs = plots_hab, ncol = 2)
```

## <span class="en">Discussion</span><span class="de">Diskussion</span>

<div class="discussion-box">

<div class="en">

**Invasion is not uniform across habitats.**

The EUNIS habitat-level analysis reveals that alien species invasion is highly heterogeneous. Some habitats show alien proportions exceeding 10%, while others remain virtually alien-free. This heterogeneity is not random: it follows clear ecological rules.

**Which habitats are most invaded and why?**

The most invaded habitats in Austria share several characteristics:

- **High disturbance frequency.** Habitats subject to regular physical disturbance (ploughing, mowing, trampling, flooding) create open niches that aliens can colonize. Agricultural fields, ruderal sites, and urban wastelands experience repeated disturbance cycles that prevent competitive exclusion by established natives.

- **High resource availability.** Nutrient-rich habitats (fertilised grasslands, floodplains with sediment deposition, roadside verges with runoff accumulation) provide the resources needed for rapid alien establishment. The fluctuating resource hypothesis (Davis et al. 2000) predicts that invasion peaks during resource pulses, and these habitats experience such pulses regularly.

- **Strong human connectivity.** Habitats along transport corridors (railway embankments, highway margins, canal banks) receive a continuous supply of alien propagules. The combination of disturbance, propagule pressure, and resource availability makes these habitats ideal invasion gateways.

**Which habitats resist invasion?**

The least invaded habitats include:

- **Alpine and subalpine grasslands.** Harsh climatic conditions (short growing seasons, frost, high UV radiation, thin soils) exclude most alien species, which tend to be adapted to temperate lowland conditions. The few aliens that reach alpine elevations are typically cold-tolerant ruderals that remain confined to disturbed microsites (ski areas, mountain huts, construction sites).

- **Oligotrophic bogs and fens.** Nutrient-poor, waterlogged conditions create strong environmental filters. The specialised native flora (Sphagnum mosses, carnivorous plants, sedges) is adapted to these extreme conditions; most alien species cannot tolerate them.

- **Closed-canopy forests.** Light limitation is a powerful filter against alien establishment. Dense spruce forests or beech forests with closed canopies receive very little light at ground level, preventing alien germination. However, forest edges, gaps, and managed forests with thinning operations show higher alien presence.

**The 100-fold variation:**

Chytry et al. (2008) demonstrated that European habitats vary 100-fold in their susceptibility to alien plant invasions. Our Austrian data confirms this continental pattern at a national scale. The most invaded habitat type can harbour 50--100 times more aliens (proportionally) than the least invaded type. This variation is far larger than differences between countries or biogeographic regions, indicating that **habitat type is the primary determinant of invasion susceptibility**, more important than climate, geography, or even propagule pressure.

**Accumulation curve shapes differ between habitats:**

The within-habitat native vs. alien accumulation curves reveal additional detail:

- In heavily invaded habitats, the alien curve shape approaches the native curve shape, indicating that aliens have spread broadly within those habitat types and achieved a relatively even spatial distribution. These aliens behave almost like "honorary natives" within their preferred habitats.

- In lightly invaded habitats, the alien curve is steeper and more erratic, reflecting a few spatially clustered occurrences rather than widespread presence. This is the signature of early-stage invasion: a species arrives at one or two sites but has not yet had time (or conditions) to spread.

**Practical application for conservation:**

This habitat-level analysis provides a prioritisation framework for invasion management:

1. **High priority:** Monitor and manage the most invaded habitat types, where aliens are already abundant and spreading. Early intervention here can prevent further entrenchment.
2. **Moderate priority:** Watch habitats at the transition zone, where aliens are present but not yet dominant. These are the "invasion frontiers" where management can be most cost-effective.
3. **Lower priority:** Alpine and oligotrophic habitats currently resist invasion. However, climate change may erode this resistance by warming alpine conditions and altering hydrological regimes in wetlands. Long-term monitoring is still warranted.

</div>

<div class="de">

**Invasion ist nicht gleichmäßig über Habitate verteilt.**

Die EUNIS-Habitat-Analyse zeigt, dass die Invasion durch Alien-Arten sehr heterogen ist. Einige Habitate zeigen Alien-Anteile über 10%, während andere praktisch alien-frei bleiben. Diese Heterogenität folgt klaren ökologischen Regeln.

**Welche Habitate sind am stärksten invadiert und warum?**

Die am stärksten invadierten Habitate in Österreich teilen mehrere Merkmale:

- **Hohe Störungsfrequenz.** Regelmäßig physisch gestörte Habitate (Pflügen, Mähen, Trampeln, Überflutung) schaffen offene Nischen für Alien-Besiedelung.

- **Hohe Ressourcenverfügbarkeit.** Nährstoffreiche Habitate (gedüngte Grasländer, Flussauen mit Sedimentablagerung) liefern die Ressourcen für schnelle Alien-Etablierung. Die Hypothese der schwankenden Ressourcen (Davis et al. 2000) sagt voraus, dass Invasion während Ressourcenpulsen ihren Höhepunkt erreicht.

- **Starke menschliche Konnektivität.** Habitate entlang von Transportkorridoren erhalten eine kontinuierliche Versorgung mit Alien-Propagules.

**Welche Habitate widerstehen der Invasion?**

- **Alpine und subalpine Grasländer.** Raue Klimabedingungen schließen die meisten Alien-Arten aus.
- **Oligotrophe Moore und Feuchtgebiete.** Nährstoffarme, wassergesättigte Bedingungen schaffen starke Umweltfilter.
- **Geschlossene Wälder.** Lichtlimitierung verhindert Alien-Etablierung unter dichtem Kronendach.

**Die 100-fache Variation:**

Chytry et al. (2008) zeigten, dass europäische Habitate sich 100-fach in ihrer Anfälligkeit für Pflanzeninvasionen unterscheiden. Unsere österreichischen Daten bestätigen dies: **Habitattyp ist der primäre Determinant der Invasionsanfälligkeit**, wichtiger als Klima, Geografie oder Propagule Pressure.

**Akkumulationskurvenformen unterscheiden sich:**

- In stark invadierten Habitaten nähert sich die Alien-Kurvenform der heimischen an — Aliens haben sich breit verteilt.
- In schwach invadierten Habitaten ist die Alien-Kurve steiler und unregelmäßiger — das Zeichen einer Frühphasen-Invasion.

**Praktische Anwendung für den Naturschutz:**

1. **Hohe Priorität:** Am stärksten invadierte Habitattypen überwachen und managen.
2. **Mittlere Priorität:** Habitate an der Übergangszone beobachten — die "Invasionsfronten", wo Management am kosteneffektivsten sein kann.
3. **Niedrigere Priorität:** Alpine und oligotrophe Habitate, die derzeit Invasion widerstehen, langfristig monitoren, besonders angesichts des Klimawandels.

</div>

</div>

---

# <span class="en">Synthesis: Five Stories, One Landscape</span><span class="de">Synthese: Fünf Geschichten, eine Landschaft</span>

<div class="results-box">

<div class="en">

These five findings form a coherent narrative about Austrian plant diversity and its invasion by alien species:

**1. Geography sets the stage.** The hotspot maps reveal that native and alien species richness follow fundamentally different spatial rules. Natives peak in the topographically complex pre-Alpine zones; aliens concentrate around human infrastructure in the lowlands. This spatial decoupling is the geographic expression of two different species pools shaped by two different sets of processes.

**2. Spatial accumulation confirms the difference.** The normalized accumulation curves demonstrate that aliens have higher spatial beta diversity than natives, requiring more landscape coverage to detect the same proportion of species. This is the quantitative confirmation of what the maps show qualitatively: aliens are spatially heterogeneous in composition.

**3. The story is not over.** The Michaelis-Menten extrapolation reveals that we know only ~79% of the alien flora, compared to ~94% of natives. The 21% gap represents an invasion debt — aliens already present but not yet detected, or species en route through ongoing trade and climate change. Austria's alien flora will continue to grow.

**4. Elevation drives diversity.** The dominance of turnover over nestedness reveals that Austria's high beta diversity is generated by genuine species replacement along the elevation gradient, not by differential species loss. This has direct consequences for conservation: no single reserve can capture the full flora; a distributed network is essential.

**5. Habitat filters invasion.** Alien invasion is not a uniform process. It concentrates in disturbed, nutrient-rich, well-connected habitats and is excluded from harsh, oligotrophic, and closed-canopy environments. This habitat-level heterogeneity provides actionable targets for conservation management.

Together, these findings demonstrate that **spatial species accumulation curves** are far more than a sampling methodology. They are a lens through which the interplay of biogeography, invasion ecology, conservation biology, and environmental filtering becomes visible. The `spacc` package makes this lens accessible for any community dataset with geographic coordinates.

</div>

<div class="de">

Diese fünf Ergebnisse bilden eine kohärente Geschichte über Österreichs Pflanzendiversität und ihre Invasion durch Alien-Arten:

**1. Geografie setzt die Bühne.** Die Hotspot-Karten zeigen, dass heimischer und Alien-Artenreichtum grundlegend verschiedenen räumlichen Regeln folgen. Einheimische erreichen ihren Höhepunkt in den topografisch komplexen Voralpen; Aliens konzentrieren sich um menschliche Infrastruktur im Tiefland.

**2. Räumliche Akkumulation bestätigt den Unterschied.** Die normalisierten Kurven zeigen, dass Aliens eine höhere räumliche Beta-Diversität haben als Einheimische. Dies ist die quantitative Bestätigung dessen, was die Karten qualitativ zeigen.

**3. Die Geschichte ist nicht zu Ende.** Die Extrapolation zeigt, dass wir nur ~79% der Alien-Flora kennen gegenüber ~94% der Einheimischen. Die 21%-Lücke repräsentiert eine Invasionsschuld.

**4. Höhenlage treibt Diversität.** Die Dominanz von Turnover über Nestedness zeigt, dass Österreichs hohe Beta-Diversität durch echten Artenersatz entlang des Höhengradienten erzeugt wird. Kein einzelnes Reservat kann die gesamte Flora erfassen.

**5. Habitat filtert Invasion.** Alien-Invasion konzentriert sich in gestörten, nährstoffreichen Habitaten und wird aus rauen, oligotrophen Umgebungen ausgeschlossen. Diese Heterogenität liefert umsetzbare Ziele für das Naturschutzmanagement.

Zusammen zeigen diese Ergebnisse, dass **räumliche Artenakkumulationskurven** weit mehr als eine Beprobungsmethode sind. Sie sind eine Linse, durch die das Zusammenspiel von Biogeografie, Invasionsökologie, Naturschutzbiologie und Umweltfilterung sichtbar wird.

</div>

</div>

---

## <span class="en">References</span><span class="de">Literaturverzeichnis</span>

<div class="lit-box">

- Baselga, A. (2010). Partitioning the turnover and nestedness components of beta diversity. *Global Ecology and Biogeography*, 19, 134--143.
- Chao, A. & Jost, L. (2012). Coverage-based rarefaction and extrapolation. *Methods in Ecology and Evolution*, 3, 217--226.
- Chytry, M. et al. (2008). Habitat invasions by alien plants: a quantitative comparison among Mediterranean, subcontinental and oceanic regions of Europe. *Journal of Applied Ecology*, 45, 448--458.
- Chytry, M. et al. (2009). European map of alien plant invasions based on the quantitative assessment across habitats. *Diversity and Distributions*, 15, 98--107.
- Chytry, M. et al. (2016). European Vegetation Archive (EVA): an integrated database of European vegetation plots. *Applied Vegetation Science*, 19, 173--180.
- Colwell, R.K. & Coddington, J.A. (1994). Estimating terrestrial biodiversity through extrapolation. *Philosophical Transactions of the Royal Society B*, 345, 101--118.
- Davis, M.A. et al. (2000). Fluctuating resources in plant communities: a general theory of invasibility. *Journal of Ecology*, 88, 528--534.
- Essl, F. et al. (2011). Socioeconomic legacy yields an invasion debt. *Proceedings of the National Academy of Sciences*, 108, 203--207.
- Grabherr, G. et al. (2003). *Die Pflanzengesellschaften Osterreichs*. Gustav Fischer Verlag.
- Kier, G. et al. (2009). A global assessment of endemism and species richness across island and mainland regions. *PNAS*, 106, 9322--9327.
- Lockwood, J.L. et al. (2005). The role of propagule pressure in explaining species invasions. *Trends in Ecology & Evolution*, 20, 223--228.
- Partel, M. et al. (2011). Dark diversity: shedding light on absent species. *Trends in Ecology & Evolution*, 26, 596--600.
- Pysek, P. et al. (2010). Disentangling the role of environmental and human pressures on biological invasions across Europe. *PNAS*, 107, 12157--12162.
- Rahbek, C. (1995). The elevational gradient of species richness: a uniform pattern? *Ecography*, 18, 200--205.
- Seebens, H. et al. (2017). No saturation in the accumulation of alien species worldwide. *Nature Communications*, 8, 14435.
- Socolar, J.B. et al. (2016). How should beta-diversity inform biodiversity conservation? *Trends in Ecology & Evolution*, 31, 67--80.
- Tjorve, E. (2010). How to resolve the SLOSS debate: lessons from species-diversity models. *Journal of Theoretical Biology*, 264, 604--612.
- Wilson, J.R.U. et al. (2007). Residence time and potential range: crucial considerations in modelling plant invasions. *Diversity and Distributions*, 13, 11--22.

</div>

---

## <span class="en">Session Info</span><span class="de">Session Info</span>

```{r session-info}
sessionInfo()
```
